---
title: "MLB EDA Project"
author: "Amor Ai"
date: '2022-06-16'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load the data}
library(tidyverse)
batted_data <- read_csv("http://www.stat.cmu.edu/cmsac/sure/2022/materials/data/sports/eda_projects/mlb_batted_balls_2022.csv")

Batting <- as_tibble(batted_data)
```

# General Overview of Dataset/Explanation of Data Structure

```{r overview}
dim(Batting) #dimensions
colnames(Batting)
# head(Batting)
# tail(Batting)
# class(Batting)

summary(Batting)
```

7085 observations of all batted balls from the current 2022 MLB season, through June 10th, courtesy of baseballsavant.com and accessed using the baseballr package

31 variables


# Hypotheses

1. In what ways are players getting out? (1D - categorical)
2. How does release speed affect launch speed? (1D - continuous)
3. How do left-handed batters and right-handed batters differ? (2D)
    a.  Does the event differ according to the handedness of the batter? 
    b.  Does hit location depend on the handedness of the batter?
4. How does the probability of pitch type thrown change depending on the count? (2D)


## What is the most common way of getting out? 

```{r getting out}
#table(Batting$events)

#Filter events to just outs:
Batting_outs <- Batting %>% 
  filter(events != "double", events != "field_error", 
         events != "fielders_choice", events !="home_run", 
         events !="single", events !="triple") %>% 
  mutate(events = 
           fct_recode(events, 
                      "Double Play" = "double_play",
                      "Field Out" = "field_out",
                      "Fielder's Choice Out" = "fielders_choice_out",
                      "Force Out" = "force_out",
                      "Grounded into Double Play" = "grounded_into_double_play",
                      "Sacrifice Bunt" = "sac_bunt",
                      "Sacrifice Fly" = "sac_fly",
                      "Sacrifice Fly Double Play" = "sac_fly_double_play")) 

#Table of outs
table(Batting_outs$events)
sort(table(Batting_outs$events), decreasing = TRUE)

#Sort as decreasing
Batting_outs$events <- factor(Batting_outs$events, names(sort(table(Batting_outs$events), increasing = TRUE)))

#Graph
Batting_outs %>% 
  ggplot(aes(x = events)) +
  geom_bar(fill= "#238A8DFF") +
  coord_flip() +
  labs(y= "count", x = "Type of outs", title = "Distribution of ways of getting out") +
  ggthemes::scale_color_colorblind() +
  theme(axis.text = element_text(angle = 90)) +
  theme_bw()
  
```

## How does release speed affect launch speed?

```{r release and launch speed}
#Remove NA's
Batting <- Batting %>% 
  filter(!is.na(release_speed), !is.na(launch_speed)) 

Batting %>% 
  ggplot(aes(x = release_speed, y = launch_speed)) +
  geom_point(alpha = 0.5, color= "#238A8DFF") +
  theme_bw() +
  labs(y= "Launch Speed (mph)", x = "Release Speed (mph)", title = "Comparing pitcher's release speed to batter's launch speed")
```

## Does the event differ according to the handedness of the batter?

```{r proportion of on-base vs out vs home run based on hand}

# Change labels for events
table(Batting$events)

filter_Batting <- Batting %>% 
  filter(!is.na(events)) %>% 
  mutate(events = 
           fct_recode(events, 
                      "On-base" = "double",
                      "On-base" = "field_error",
                      "Out" = "double_play",
                      "Out" = "field_out",
                      "On-base" = "fielders_choice",
                      "Out" = "fielders_choice_out",
                      "Homerun" = "home_run",
                      "Out" = "force_out",
                      "Out" = "grounded_into_double_play",
                      "Out" = "sac_bunt",
                      "Out" = "sac_fly",
                      "Out" = "sac_fly_double_play",
                      "On-base" = "single",
                      "On-base" = "triple"))

#table(filter_Batting$events)

left_Batting <- filter_Batting %>% 
  filter(stand == "L") %>% 
  group_by(events) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "left")

right_Batting <- filter_Batting %>% 
  filter(stand == "R") %>% 
  group_by(events) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "right")

prop_Batting <- left_Batting %>% 
  rbind(right_Batting)

prop_Batting %>% 
  ggplot(aes(x = events, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw() +
  labs(y= "Proportion", x = "Events", title = "Distribution of Events Based on Batting Left-handed vs Right-handed") 


```
### Is the difference significance?

```{r test for significance}

#Proportion of on-base

prop_Batting %>% 
  filter(events == "On-base") %>% 
  ggplot(aes(x = events, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw()

stand_onbase <- prop_Batting %>% 
  filter(events == "On-base")

t.test(stand_onbase$prop)

#Proportion of outs

prop_Batting %>% 
  filter(events == "Out") %>% 
  ggplot(aes(x = events, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw()

stand_outs <- prop_Batting %>% 
  filter(events == "Out")

t.test(stand_outs$prop)

#Proportion of home-runs

prop_Batting %>% 
  filter(events == "Homerun") %>% 
  ggplot(aes(x = events, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw()

stand_hr <- prop_Batting %>% 
  filter(events == "Homerun")

t.test(stand_hr$prop)
```

***YES, at alpha level 0.05, we have sufficient evidence to reject the null hypothesis that the proportion of on-base, outs, and home-runs between left and right-handed batters are different.***

### Does hit location depend on the handedness of the batter?

```{r hit location depending on handedness}

filter_location <- Batting %>% 
  filter(!is.na(hit_location)) %>%  #get rid of NA's
  group_by(hit_location, stand) 

left_location <- filter_location %>% 
  filter(stand == "L") %>% 
  group_by(hit_location) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "left")

right_location <- filter_location %>% 
  filter(stand == "R") %>% 
  group_by(hit_location) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "right")

prop_location <- left_location %>% 
  rbind(right_location)

prop_location %>% 
  ggplot(aes(x = hit_location, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw() +
  labs(y= "Proportion", x = "Hit Location", title = "Distribution of Hit Location Based on Batting Left-handed vs Right-handed") +
  scale_x_discrete(limits = c(1:9))

```
```{r hit location proportion overlap on actual baseball field?}

```

## How likely is a pitch type for each count?

```{r load new variable}

# overall pitch types
edit_pitch <- Batting %>% 
  filter(!is.na(pitch_type)) %>% 
  mutate(pitch_type = fct_recode(pitch_type, "Changeup" = "CH", 
                                 "Breaking ball" = "CU",
                                 "Changeup" = "EP",
                                 "Fastball" = "FA",
                                 "Fastball" = "FC", 
                                 "Fastball" = "FF", 
                                 "Fastball" = "FS",
                                 "Breaking ball" = "KC",  
                                 "Fastball" = "SI",  
                                 "Breaking ball" = "SL"))

# 0-0
zero_zero <- edit_pitch %>% 
  filter(balls == 0) %>% 
  filter(strikes == 0) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "zero_zero")


# 0-1
zero_one <- edit_pitch %>% 
  filter(balls == 0) %>% 
  filter(strikes == 1) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "zero_one")

# 0-2
zero_two <- edit_pitch %>% 
  filter(balls == 0) %>% 
  filter(strikes == 2) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "zero_two")

# 1-0
one_zero <- edit_pitch %>% 
  filter(balls == 1) %>% 
  filter(strikes == 0) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "one_zero")

# 1-1
one_one <- edit_pitch %>% 
  filter(balls == 1) %>% 
  filter(strikes == 1) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "one_one")

# 1-2
one_two <- edit_pitch %>% 
  filter(balls == 1) %>% 
  filter(strikes == 2) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "one_two")

# 2-0
two_zero <- edit_pitch %>% 
  filter(balls == 2) %>% 
  filter(strikes == 0) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "two_zero")


# 2-1
two_one <- edit_pitch %>% 
  filter(balls == 2) %>% 
  filter(strikes == 1) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "two_one")

# 2-2
two_two <- edit_pitch %>% 
  filter(balls == 2) %>% 
  filter(strikes == 2) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "two_two")

# 3-0
three_zero <- edit_pitch %>% 
  filter(balls == 3) %>% 
  filter(strikes == 0) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "three_zero")

# 3-1
three_one <- edit_pitch %>% 
  filter(balls == 3) %>% 
  filter(strikes == 1) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "three_one")

# 3-2
three_two <- edit_pitch %>% 
  filter(balls == 3) %>% 
  filter(strikes == 2) %>% 
  group_by(pitch_type) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n), count = "three_two")

prop_count <- zero_zero %>% 
  rbind(zero_one, zero_two, one_zero, one_one, one_two, two_one, two_two, 
        two_zero, three_zero, three_one, three_two)

prop_count %>% 
  ggplot(aes(x = pitch_type, y = prop, fill = count)) +
  geom_col(position = "dodge") +
  theme_bw() 

# write csv so all objects don't have to be loaded every time
write_csv(prop_count, "prop_count.csv")

# load csv
library(readr)
prop_count <- read_csv("prop_count.csv")
View(prop_count)

```

```{r pitch type}

changeup_props <- prop_count %>% 
  filter(pitch_type == "Changeup") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))

breaking_props <- prop_count %>% 
  filter(pitch_type == "Breaking ball") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))
  
  
breaking_props <- breaking_props %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))

fast_props <- prop_count %>% 
  filter(pitch_type == "Fastball")
fast_props <- prop_count %>% 
  filter(pitch_type == "Fastball") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-0" = "three_zero", 
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))


# checking to make sure 3-0 is only fastballs
try <- Batting %>% 
  filter(balls == 3) %>% 
  filter(strikes == 0) %>% 
  arrange(pitch_type)


counts <- c("0-0", "0-1", "0-2", "1-0", "1-1", "1-2", "2-0", "2-1", "2-2",
            "3-0", "3-1", "3-2")
counts2 <- c("0-0", "0-1", "0-2", "1-0", "1-1", "1-2", "2-0", "2-1", "2-2",
            "3-1", "3-2")

changeup_props %>% 
  ggplot(aes(x = count, y = prop)) +
  #geom_point() +
  theme_bw() +
  geom_col() +
  labs(title = "Changeup") +
  scale_x_discrete(limits = counts2)

breaking_props %>% 
  ggplot(aes(x = count, y = prop)) +
  theme_bw() +
  geom_col() +
  labs(title = "Breaking balls") +
  scale_x_discrete(limits = counts2)

fast_props %>% 
  ggplot(aes(x = count, y = prop)) +
  theme_bw() +
  geom_col() +
  labs(title = "Fastball") +
  scale_x_discrete(limits = counts)


```


# Initial Cluster 1: 
## Clustering batted balls based on hit distance and launch speed

```{r clean data}
Batting <- Batting %>% 
  filter(!is.na(hit_distance_sc), !is.na(launch_speed)) # Look at data/ remove na's

# Standardize first

std_batting <- Batting %>%
  mutate(std_hit = as.numeric(scale(hit_distance_sc)),
         std_speed = as.numeric(scale(launch_speed)))

std_batting %>%
  ggplot(aes(x = std_hit, y = std_speed)) +
  geom_point(alpha = 0.5) + 
  theme_bw() +
  coord_fixed()
```


### Trying kmeans

```{r k-means}

# Initialize number of clusters to search over
n_clusters_search <- 2:12
tibble(total_wss = 
         # Compute total WSS for each number by looping with sapply
         sapply(n_clusters_search,
                function(k) {
                  kmeans_results <- kmeans(dplyr::select(std_batting,
                                                         std_hit,
                                                         std_speed),
                                           centers = k, nstart = 30)
                  # Return the total WSS for choice of k
                  return(kmeans_results$tot.withinss)
                })) %>%
  mutate(k = n_clusters_search) %>%
  ggplot(aes(x = k, y = total_wss)) +
  geom_line() + geom_point() +
  labs(x = "Number of clusters K", y = "Total WSS") +
  theme_bw()

## k-means

init_kmeans <- 
  kmeans(dplyr::select(std_batting,
                       std_hit, std_speed),
         algorithm = "Lloyd", centers = 3,
         nstart = 30)

std_batting %>%
  mutate(batting_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = std_hit, y = std_speed,
             color = batting_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

# k-means++

library(flexclust)

init_kmeanspp <- 
  kcca(dplyr::select(std_batting,
                     std_hit, std_speed), k = 3,
       control = list(initcent = "kmeanspp"))
std_batting %>%
  mutate(batting_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = std_hit, y = std_speed,
             color = batting_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

```

### Trying hclust
```{r hclust}

#minimax

library(protoclust)

batted_dist <- dist(dplyr::select(std_batting,
                                  std_hit, std_speed))

batted_minimax <- protoclust(batted_dist)

# Look at dendrogram
library(ggdendro)
ggdendrogram(batted_minimax, 
             theme_dendro = FALSE, 
             labels = FALSE, 
             leaf_labels = FALSE) + 
  labs(y = "Maximum dissimilarity from prototype") +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())


minimax_batted_clusters <- 
  protocut(batted_minimax, k = 3) #where to cut tree

std_batting %>%
  mutate(batted_clusters = 
           as.factor(minimax_batted_clusters$cl)) %>%
  ggplot(aes(x = std_hit, y = std_speed,
             color = batted_clusters)) +
  geom_point(alpha = 0.3) + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")


#single

batted_dist <- dist(dplyr::select(std_batting,
                                  std_hit, std_speed))

batted_single_hclust <-
  hclust(batted_dist, method = "single")


std_batting %>%
  mutate(batted_clusters = 
           as.factor(cutree(batted_complete_hclust, k=3))) %>% #cut dendrogram using chosen height h
  ggplot(aes(x = std_hit, y = std_speed,
             color = batted_clusters)) +
  geom_point(alpha = 0.5) + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")

#complete

batted_complete_hclust <-
  hclust(batted_dist, method = "complete")

std_batting %>%
  mutate(batted_clusters = 
           as.factor(cutree(batted_complete_hclust, k=3))) %>% #cut dendrogram using chosen height h
  ggplot(aes(x = std_hit, y = std_speed,
             color = batted_clusters)) +
  geom_point(alpha = 0.5) + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")


# How does events relate to our clustering results?

table("Clusters" = minimax_batted_clusters$cl,
      "Events" = filter_Batting$events) 

mosaicplot(table(minimax_batted_clusters$cl,
      filter_Batting$events), main = "Relationship between clusters and events")

```

# Initial Cluster 2: 
## Clustering players based on hit distance and launch speed (to see which players are strong hitters)

```{r cluster players}
# Summarize batted balls by player name

#mean(table(Batting$player_name)) 
#average at bats = 17, remove players who've batted less than 5 times 

players <- Batting %>% 
  group_by(player_name) %>% 
  filter(n() > 5) %>%  #filter players who've batted more than 5 times
  summarise_all(.funs = mean, .groups = "drop") #take player averages

filter_players <- players %>% 
  drop_na(hit_distance_sc, launch_speed) #get rid of rows that have NA's in hit distance and launch speed


# Try hierarchical clustering
#minimax

library(protoclust)
player_dist <- dist(dplyr::select(filter_players,
                                  hit_distance_sc, launch_speed))
player_minimax <- protoclust(player_dist)

# Look at dendrogram
library(ggdendro)
ggdendrogram(player_minimax, 
             theme_dendro = FALSE, 
             labels = FALSE, 
             leaf_labels = FALSE) + 
  labs(y = "Maximum dissimilarity from prototype") +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())


minimax_player_clusters <- 
  protocut(player_minimax, k = 3) #where to cut tree

filter_players %>%
  mutate(player_clusters = 
           as.factor(minimax_player_clusters$cl)) %>%
  ggplot(aes(x = hit_distance_sc, y = launch_speed,
             color = player_clusters)) +
  geom_point(alpha = 0.5) + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")

# View who the prototypes are
player_prototypes <- filter_players %>%
  dplyr::select(player_name, hit_distance_sc, launch_speed) %>%
  slice(minimax_player_clusters$protos)

# Label prototypes on graph 
filter_players %>%
  mutate(player_clusters = 
           as.factor(minimax_player_clusters$cl)) %>%
  ggplot(aes(x = hit_distance_sc, y = launch_speed,
             color = player_clusters)) +
  geom_point(alpha = 0.5) + 
  geom_label(data = mutate(player_prototypes, 
                           player_clusters = 
                             as.factor(c(1,2,3))), aes(label = player_name)) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")

# View cluster 3

best_batters <- filter_players %>%
  mutate(player_clusters = 
           as.factor(minimax_player_clusters$cl)) %>% 
  filter(player_clusters == 3) %>% 
  pull(player_name)

data.frame(best_batters) #group of strongest batters this season
```



# Discussion

- point out limitations & direction for next steps (future work)

## References










