---
title: "Exploratory Data Analysis on MLB Batted Ball Data"
author: "Amor Ai & Jackie Jovanovic"
date: "June 24th, 2022"
output:
  xaringan::moon_reader:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r, library, message = FALSE, warning = FALSE, include = FALSE}
library(tidyverse)
library(patchwork)

batted_data <- read_csv("http://www.stat.cmu.edu/cmsac/sure/2022/materials/data/sports/eda_projects/mlb_batted_balls_2022.csv")

Batting <- as_tibble(batted_data)
```


# Data

- Batted balls data from the 2022 MLB season, through June 10th

- There are 7,085 observations and 31 variables

- Information on batter's stand, before and after counts, speed, batted ball type, hit distance, pitch type, outcome, and score

```{r, table, echo = FALSE}
Batting %>% 
  select(player_name, stand, events, release_speed, bb_type, 
         hit_distance_sc) %>% 
  head(5) %>% 
  knitr::kable(format = 'html')
```

---

# In what ways are players gettting out?
```{r, getting_out, echo = FALSE, message=FALSE, warning=FALSE, fig.align='center'}
#Filter events to just outs:
Batting_outs <- Batting %>% 
  filter(events != "double", events != "field_error", 
         events != "fielders_choice", events !="home_run", 
         events !="single", events !="triple") %>% 
  mutate(events = 
           fct_recode(events, 
                      "Double Play" = "double_play",
                      "Field Out" = "field_out",
                      "Fielder's Choice Out" = "fielders_choice_out",
                      "Force Out" = "force_out",
                      "Grounded into Double Play" = "grounded_into_double_play",
                      "Sacrifice Bunt" = "sac_bunt",
                      "Sacrifice Fly" = "sac_fly",
                      "Sacrifice Fly Double Play" = "sac_fly_double_play"))

#Sort as decreasing
Batting_outs$events <- factor(Batting_outs$events,
                              names(sort(table(Batting_outs$events),
                                         increasing = TRUE)))
#Graph
Batting_outs %>% 
  ggplot(aes(x = events)) +
  geom_bar(fill= "#238A8DFF") +
  coord_flip() +
  labs(y= "count", x = "Type of outs", 
       title = "Distribution of ways of getting out") +
  ggthemes::scale_color_colorblind() +
  theme(axis.text = element_text(angle = 90)) +
  theme_bw()
```


---

# How does release speed affect launch speed?
```{r, speed, echo=FALSE, fig.align='center', out.width='60%'}
#Remove NA's
Batting <- Batting %>% 
  filter(!is.na(release_speed), !is.na(launch_speed)) 
Batting %>% 
  ggplot(aes(x = release_speed, y = launch_speed)) +
  geom_point(alpha = 0.5, color= "#238A8DFF") +
  theme_bw() +
  labs(y= "Launch Speed (mph)", x = "Release Speed (mph)", title = "Comparing pitcher's release speed to batter's launch speed")
```

---

# How do left-handed batters and right-handed batters differ?

1.  Does the event differ according to the handedness of the batter?
```{r, hand_props, echo=FALSE, message = FALSE, fig.align='center', out.width='60%'}
# Change labels for events
filter_Batting <- Batting %>% 
  filter(!is.na(events)) %>% 
  mutate(events = 
           fct_recode(events, 
                      "On-base" = "double",
                      "On-base" = "field_error",
                      "Out" = "double_play",
                      "Out" = "field_out",
                      "On-base" = "fielders_choice",
                      "Out" = "fielders_choice_out",
                      "Homerun" = "home_run",
                      "Out" = "force_out",
                      "Out" = "grounded_into_double_play",
                      "Out" = "sac_bunt",
                      "Out" = "sac_fly",
                      "Out" = "sac_fly_double_play",
                      "On-base" = "single",
                      "On-base" = "triple"))

left_Batting <- filter_Batting %>% 
  filter(stand == "L") %>% 
  group_by(events) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "left")
right_Batting <- filter_Batting %>% 
  filter(stand == "R") %>% 
  group_by(events) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n), hand = "right")
prop_Batting <- left_Batting %>% 
  rbind(right_Batting)
prop_Batting %>% 
  ggplot(aes(x = events, y = prop, fill = hand)) +
  geom_col(position = "dodge") +
  theme_bw() +
  labs(y= "Proportion", x = "Events", title = "Distribution of Events Based on Batting Left-handed vs Right-handed") 
```


1.  Does hit location depend on the handedness of the batter?

---

# How does the probability of pitch type thrown change depending on the count?

```{r, pitches, include = FALSE}
edit_pitch <- Batting %>% 
  filter(!is.na(pitch_type)) %>% 
  mutate(pitch_type = fct_recode(pitch_type, "Changeup" = "CH", 
                                 "Breaking ball" = "CU",
                                 "Changeup" = "EP",
                                 "Fastball" = "FA",
                                 "Fastball" = "FC", 
                                 "Fastball" = "FF", 
                                 "Fastball" = "FS",
                                 "Breaking ball" = "KC",  
                                 "Fastball" = "SI",  
                                 "Breaking ball" = "SL"))

prop_count <- read_csv("prop_count.csv")

changeup_props <- prop_count %>% 
  filter(pitch_type == "Changeup") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))

breaking_props <- prop_count %>% 
  filter(pitch_type == "Breaking ball") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))

fast_props <- prop_count %>% 
  filter(pitch_type == "Fastball") %>% 
  mutate(count = fct_recode(count,
                            "0-0" = "zero_zero",
                            "0-1" = "zero_one",
                            "0-2" = "zero_two",
                            "1-0" = "one_zero", 
                            "1-1" = "one_one", 
                            "1-2" = "one_two",
                            "2-0" = "two_zero",
                            "2-1" = "two_one", 
                            "2-2" = "two_two",
                            "3-0" = "three_zero", 
                            "3-1" = "three_one", 
                            "3-2" = "three_two"))
```

```{r, ptype_counts, echo = FALSE, fig.align='center', fig.width=10}
counts <- c("0-0", "0-1", "0-2", "1-0", "1-1", "1-2", "2-0", "2-1", "2-2",
            "3-0", "3-1", "3-2")
counts2 <- c("0-0", "0-1", "0-2", "1-0", "1-1", "1-2", "2-0", "2-1", "2-2",
             "3-1", "3-2")


g1 <- changeup_props %>% 
  ggplot(aes(x = count, y = prop)) +
  theme_bw() +
  geom_col(fill= "#238A8DFF") +
  labs(title = "Changeup") +
  scale_x_discrete(limits = counts2)

g2 <- breaking_props %>% 
  ggplot(aes(x = count, y = prop)) +
  theme_bw() +
  geom_col(fill= "#238A8DFF") +
  labs(title = "Breaking balls") +
  scale_x_discrete(limits = counts2)

g3 <- fast_props %>% 
  ggplot(aes(x = count, y = prop)) +
  theme_bw() +
  geom_col(fill= "#238A8DFF") +
  labs(title = "Fastball") +
  scale_x_discrete(limits = counts)

g1 + g2 + g3
```

--

The only type of pitch thrown during a 3-0 count is a fastball.

---

# Conclusions

  - first
  
  - second
  
  - third
  
  - The probability of a pitch being thrown is not constant throughout an at bat
      - the only type of pitch thrown during a 3-0 count is a fastball